---
title: "Dragonfish_Final"
author: "Justin Lau, Christopher Peters, James Kim, John Santoianni"
date: "2022-12-10"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, packages, include=FALSE}
library(tidyverse)
library(Momocs)
library(knitr)
library(vroom)
library(ape)
library(phytools)
library(wesanderson)
library(RRphylo)
library(ggtree)
```

# Introduction

Stomiidae is a family of deep-sea ray-finned fish and one of the oldest living species on earth. These ancient creatures have unique features such as their jaws, eyes, and gills that make them interesting targets for evolutionary biologists looking to gain insight into key development stages throughout history. As such, our research here sought to look into how the head shape of this family has evolved over time. Previous research that has been done into the family of Stomiidae is relatively limited. However, one such study done by @schnell2008evolution had a focus on uniquely preserved as well as novel features such as the pectoral girdle and vertebral column. In the findings, it was suggested that by the nature of the larger prey Stomiidae target, these anatomical skull features have proven to be beneficial in physically allowing for the engulfing and consumption of these foodstuffs. Despite the importance of prior studies like that of @schnell2008evolution, they remain limited in their ability to directly assess skull shape variables and the ecological roles of these features over evolutionary timescales and throughout lineages. As such, the goal of our study is to assess skull shape evolution across the Stomiidae, incorporating data from 18 different species in this family. Data will be collected by characterizing the shape outline of the different specimens to answer the following 3 questions:

1. What is the evolutionary rate for skull shape?
2. Is there a significant evolutionary change of skull shape?
3. Does the evolvability of the skull shape vary significantly among our major recorded lineages of Stomiidae?

# Methods

Images of various species of Dragonfish were digitized using FIJI.

Next, a Procrustes transformation was performed in order to translate, rotate, and scale the outlines for better analysis of their shape without factoring in arbitrary skull size.

Elliptical Fourier Analysis (EFA) was performed in order to describe wing shapes.
Principle Component Analysis (PCA) was then performed in order to minimize the number of variables explaining the variance in the data set.
The functions for these analyses are included in the R package "Momocs." PCA data was visualized to determine the distribution of morphoshape of the wings of the different species as well as the amount of variance described by different variables.
To prepare the comparative analysis on the PCA data, the phylogenetic tree was loaded positioned upper left.
The file names of each species were also identified using the identifier data provided by Prof. Kenaley.

The PC data was added to the identified species data accordingly.
The phylogeny tree was modified to only include the species that were identified and had a PCA data.
From here, the dragonfish skull PC1 and PC2 were separated in different variable under the Brownian Motion.

To determine the evolutionary rate, "RRphylo" package was used to determine the rate of evolution based on skull shape and from there, any statistically significant shift in shape evolution based on the Dragon fish clades were identified.

The rate of the significant evolutionary shift was identified along with the name of clades.

The phylogenetic independent contrast was performed on the PC datas to separate the original tip data independent of each other.

The PC values of forewing and hindwing data were then compared via linear model and analyzed to see if statistically significant value existed.

# Results

## Shape Analysis

::: {style="text-align: left"}
After visualizing the skull data of the dragonfish species, the skulls were scaled so that the data was adjusted for their difference in size.
This allowed for the contrast of skull shape only.
:::

<center>

```{r, echo=FALSE}

f <- list.files("CSV Files",pattern=".txt|.csv",full.names = TRUE)
out <- read_delim(f[1],delim="\t") %>% 
  as.matrix()
#transform matrix in momocs outline
out %>% 
  list() %>% 
  Out() %>% 
  coo_flipx() %>% 
  stack()
```

***Figure 1. Visualization of Dragonfish Skull Shape using EFA***

```{r, echo=FALSE}
#make a large df with vroom
out.df <- vroom::vroom(f, id = "filename")
#add species info
out.df <- out.df %>% 
  mutate(specie=gsub("\\..+","\\1",basename(filename))) %>% 
  na.omit()
#make list
outs.l <- sapply(f,function(x) out.df %>% filter(filename==x) %>% select(X,Y) %>% as.matrix)
#extract species info
species <- gsub("\\..+","\\1",basename(names(outs.l)))
outs <-  outs.l %>% 
  Out(fac=list(specie=species)) %>% 
  coo_flipx() #visualize both wings uncorrected for size
```

<center>

```{r, echo=FALSE, results='hide'}

#Procrustes, align
skull.min <- outs %>% 
  coo_nb() %>% 
  min()
outs %>%
  coo_interpolate(skull.min) %>% 
  fgProcrustes() %>% 
  stack() #visualize forewings
```

***Figure 2. Visualization of Dragonfish Skulls using EFA after the size of the skull have been equalized***

<center>

```{r, include=FALSE, results='hide'}
outs %>%
  coo_interpolate(skull.min) %>% 
   coo_align()  %>%
  fgProcrustes() %>% 
  efourier(norm=FALSE) 
```

::: {style="text-align: left"}
The visualized skulls were then processed through PCA.
This allowed us to see the distribution of skull shapes.
:::

<center>

```{r, echo=FALSE, results='hide'}
skull.pca <- outs %>%
  coo_interpolate(skull.min) %>%
   coo_align()  %>%
  coo_slide(id=1) %>% 
  fgProcrustes() %>% 
  efourier(norm=FALSE) %>% 
  PCA()

skull.pca %>% 
  plot_PCA(title = "Dragonfish Skulls")

```

***Figure 3. PCA analysis of Dragonfish Skulls***

## Comparative Analysis

::: {style="text-align: left"}
After acquiring the PCA data, the phylogenetic tree of stomidoptera was filtered to only include the species that were included in this study.
:::

<center>

```{r, fig.show='hide', display=FALSE}

stom.tree <- ape::read.tree("Stomiidae.tre")

plot(stom.tree,cex=0.1)

```

```{r, echo=FALSE, results='hide'}
stom.tree <- ladderize(stom.tree)
plot(stom.tree,cex=0.1)

stom.tree$tip.label <- gsub("_"," ",stom.tree$tip.label)
basename(names(outs))[1:5]
stom.tree$tip.label

#
out.data <- tibble(xy.file=basename(names(outs))) %>% 
  mutate(species=gsub(".txt","",xy.file))

skull.pca2 <-  tibble(xy.file=basename(rownames(skull.pca$x)),PC1=skull.pca$x[,1],PC2=skull.pca$x[,2]) %>% 
  left_join(out.data)
```

***Figure 6. Phylogenetic Tree of Stomiidae***

## Evolutionary Rates

<center>

```{r,echo=FALSE}
drops <- stom.tree$tip.label[!stom.tree$tip.label%in%unique(out.data$species)]

stom.tree2 <- drop.tip(stom.tree,drops)

plot(stom.tree2,cex=0.5)

#PC1
skull.pc1 <- skull.pca2 %>% 
    filter(species%in% stom.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull

names(skull.pc1) <-  skull.pca2%>% 
    filter(species%in% stom.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull(species)

#PC2s
skull.pc2 <- skull.pca2 %>% 
    filter(species%in% stom.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC2=mean(PC2)) %>% 
  pull(PC2)

names(skull.pc2) <-  skull.pca2%>% 
    filter(species%in% stom.tree2$tip.label) %>% 
  group_by(species) %>%
  summarize(PC2=mean(PC2)) %>% 
  pull(species)
```

***Figure 7. Adjusted phylogenetic tree of stomiidae included in the study***

```{r, echo=FALSE}

skullPC1.BM<-brownie.lite(stom.tree2,skull.pc1*10)
skullPC2.BM<-brownie.lite(stom.tree2,skull.pc2*10)

#Evolutionary rates of dragonfish skull
skullPC1.BM$sig2.single
skullPC2.BM$sig2.single
```

## Shifts in Evolutionary Rates

::: {style="text-align: left"}
The evolutionary shift was then measured to determine if any significant shift in evolutionary rate was present according to the different clades of stomiidae under the Brownian model.
:::

<center>

```{r, include= FALSE}
skullPC1.RR <- RRphylo(tree=stom.tree2,y=skull.pc2)
skullPC1.SS<- search.shift(RR=skullPC1.RR,status.type="clade")

plot(stom.tree2)

nodelabels(node = as.numeric(rownames(skullPC1.SS$single.clades)),text = rownames(skullPC1.SS$single.clades))

skullPC1.plot <- plotShift(RR=skullPC1.RR,SS=skullPC1.SS)

skullPC1.plot$plotClades()
```

```{r, Define funtion, include= FALSE}


plot_SS <- function(tre=NULL,SS=NULL,tax=NULL){
  

  nodes <- as.numeric(rownames(SS$single.clades))
  
  pal <- wes_palette("Zissou1",n=length(nodes))
  sp <- list()
  for(i in nodes){
    sp.i <- extract.clade(tre,i)$tip.label
    
    #print(head(tax))
    sub.names <- lapply(tax,function(x) x[x%in%sp.i]) 
    
    in.clades <- lapply(sub.names,function(x) length(x)>0) 
    all.of.clade <- lapply(sub.names,function(x) all(sapply(sp.i,function(z) z%in%x))) 
    
    high.clade <- names(sub.names)[last(which(all.of.clade==T))]
    all.clades <- names(sub.names)[which(in.clades==T)]
    crown <- ""
    if(high.clade!=last(names(sub.names))) crown <- "crown-"
    
    sub.clades <- NULL
    if(length(grepl("oidea",all.clades))>0) sub.clades <- all.clades[grepl("oidea",all.clades)]

    high.clade2 <- paste0(crown,high.clade,": ",paste0(sub.clades,collapse = "+"))
    sp[[paste0(i)]] <- tibble(n=i,species=sp.i,clade=high.clade2)
    
  }

  
  d <- do.call(rbind,sp)%>% 
    rename(label=species) 
  
  d2<- d %>% rename(clade_name=clade) 
  
  p <- ggtree(tre)+ scale_y_reverse()
  
  p$data <- p$data %>% left_join(d) %>% left_join(tibble(node=nodes,SS$single.clades) %>% mutate(shift=ifelse(rate.difference>0,"+","-")))
  
  p <-  p+geom_tiplab(aes(col=clade),geom="text",size=1.2)+
    geom_cladelab(data=d2,mapping=aes(node=n,col=clade_name,label=clade_name),offset=1,size=1.5)+
    geom_hilight(data=d2,mapping = aes(node = n,fill=clade_name),alpha = 0.01)+
    scale_fill_manual(values = pal)+
    scale_color_manual(values = pal)+
    theme(legend.position = "none")+geom_nodepoint(mapping=aes(subset = shift =="-"), size=5, shape=25,fill='blue',color='blue',alpha=0.7)+
    geom_nodepoint(mapping=aes(subset = shift =="+"), size=5, shape=24, fill='red',color='red',alpha=0.7)
  p <- p+xlim(NA,6)
  res <- tibble(n=nodes,SS$single.clades) %>% left_join(d %>% select(n,clade) %>% unique)
  
  return(list(plot=p,res=res))
  
}

```

```{r, plot shift in evolutionary rates}
tax.names <- readRDS("stom_classification.RDS")

skullPC1.res <- plot_SS(stom.tree2,skullPC1.SS,tax = tax.names)

skullPC1.res$plot
skullPC1.res$res
```

***Figure 8. Significant evolutionary rate found in the Stomiidae phylogeny***

## Shape Evolution Correlation

::: {style="text-align: left"}
The PC1 and PC2 data that underwent phylogenetic independent contrast showed that the PC1 data of skull shape comparison had much more significant correlation compared to the PC2.
:::

### PC1 #need to work on this section once have the tree data

<center>

```{r, PIC for PC1, echo=FALSE}
skullPC1.pic <- pic(skull.pc1,phy = stom.tree2)

PC1.pic <- tibble(
  skull=skullPC1.pic
)

PC1.pic %>% 
  ggplot(aes(x=fore,y=hind))+geom_point()+geom_smooth(method="lm")+theme_classic()

summary(lm(hind~fore,PC1.pic))

```

***Figure 9. Phylogenetic Independent Contrast of PC1***

<center>

### PC2

```{r, PIC for PC2, echo=FALSE}
hindPC2.pic <- pic(hind.pc2,phy = stom.tree2)
forePC2.pic <- pic(fore.pc2,phy = stom.tree2)
PC2.pic <- tibble(
  hind=hindPC2.pic,
  fore=forePC2.pic
)

PC2.pic %>% 
  ggplot(aes(x=fore,y=hind))+geom_point()+geom_smooth(method="lm")+ theme_classic() #plot PIC PC2

summary(lm(hind~fore,PC2.pic)) #Find R-square

```

***Figure 10. Phylogenetic Independent Contrast of PC2***

# Discussion

Various studies have been done to analyze the macroevolutionary history of stomidopteran forewings and hindwings.
One study by @le2019adaptive links morphology with natural selection, but there is little information known about rates of evolution and correlation of wing shapes.
In this study, we aimed to answer some of these questions about evolutionary history.
We first compared the rate of wing-shape evolution between forewings and hindwings.
Evolutionary rates calculated from PC1 for forewings and hindwings differed by 35.56%, whereas evolutionary rates calculated from PC2 differed by 87.11%, indicating a significant shift in evolutionary rates between forewing and hindwing.
Because forewing and hindwing were found to have different evolutionary rates, we examined shift in evolutionary rates of major lineages as it pertains to both hind- and forewing PC1.
Results obtained from both agreed with each other, in that the clade Macroheterocera: Noctuoidea underwent a downward shift in wing-shape evolutionary rate.
Next, to study fore- and hindwing shape correlation, we examined PC1 and PC2 transformed with phylogenetic information for both.
PIC analysis on fore- and hindwing PC1 showed significant correlation (R-squared: 0.4476), whereas PIC analysis using PC2 found no significant correlation (R-squared: 0.00408).

# Author Contributions

Justin - File Formatting, Code, Methods

James -

Chris -

John -

# References
