---
title: "Dragonfish_Final"
author: "Justin Lau, Christopher Peters, James Kim, John Santoianni"
date: "2022-12-10"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, packages, include=FALSE}
library(tidyverse)
library(Momocs)
library(knitr)
library(vroom)
library(ape)
library(phytools)
library(wesanderson)
library(RRphylo)
library(ggtree)
```

# Introduction

# Methods

Images of various species of Dragonfish were digitized using FIJI.

Next, a Procrustes transformation was performed in order to translate, rotate, and scale the outlines for better analysis of their shape without factoring in arbitrary skull size.

Elliptical Fourier Analysis (EFA) was performed in order to describe wing shapes.
Principle Component Analysis (PCA) was then performed in order to minimize the number of variables explaining the variance in the data set.
The functions for these analyses are included in the R package "Momocs." PCA data was visualized to determine the distribution of morphoshape of the wings of the different species as well as the amount of variance described by different variables.
To prepare the comparative analysis on the PCA data, the phylogenetic tree was loaded positioned upper left.
The file names of each species were also identified using the identifier data provided by Prof. Kenaley.

The PC data was added to the identified species data accordingly.
The phylogeny tree was modified to only include the species that were identified and had a PCA data.
From here, the dragonfish skull PC1 and PC2 were separated in different variable under the Brownian Motion.

To determine the evolutionary rate, "RRphylo" package was used to determine the rate of evolution based on skull shape and from there, any statistically significant shift in shape evolution based on the Dragon fish clades were identified.

The rate of the significant evolutionary shift was identified along with the name of clades.

The phylogenetic independent contrast was performed on the PC datas to separate the original tip data independent of each other.

The PC values of forewing and hindwing data were then compared via linear model and analyzed to see if statistically significant value existed.

# Results

## Shape Analysis

::: {style="text-align: left"}
After visualizing the skull data of the dragonfish species, the skulls were scaled so that the data was adjusted for their difference in size.
This allowed for the contrast of skull shape only.
:::

<center>

```{r, echo=FALSE}

f <- list.files("CSV Files",pattern=".txt|.csv",full.names = TRUE)
out <- read_delim(f[1],delim="\t") %>% 
  as.matrix()
#transform matrix in momocs outline
out %>% 
  list() %>% 
  Out() %>% 
  coo_flipx() %>% 
  stack()
```

***Figure 1. Visualization of Dragonfish Skull Shape using EFA***

```{r, echo=FALSE}
#make a large df with vroom
out.df <- vroom::vroom(f, id = "filename")
#add species info
out.df <- out.df %>% 
  mutate(specie=gsub("\\..+","\\1",basename(filename))) %>% 
  na.omit()
#make list
outs.l <- sapply(f,function(x) out.df %>% filter(filename==x) %>% select(X,Y) %>% as.matrix)
#extract species info
species <- gsub("\\..+","\\1",basename(names(outs.l)))
outs <-  outs.l %>% 
  Out(fac=list(specie=species)) %>% 
  coo_flipx() #visualize both wings uncorrected for size
```

<center>

```{r, echo=FALSE, results='hide'}

#Procrustes, align
skull.min <- outs %>% 
  coo_nb() %>% 
  min()
outs %>%
  coo_interpolate(skull.min) %>% 
  fgProcrustes() %>% 
  stack() #visualize forewings
```

***Figure 2. Visualization of Dragonfish Skulls using EFA after the size of the skull have been equalized***

<center>

```{r, include=FALSE, results='hide'}
outs %>%
  coo_interpolate(skull.min) %>% 
   coo_align()  %>%
  fgProcrustes() %>% 
  efourier(norm=FALSE) 
```

::: {style="text-align: left"}
The visualized skulls were then processed through PCA.
This allowed us to see the distribution of skull shapes.
:::

<center>

```{r, echo=FALSE, results='hide'}
skull.pca <- outs %>%
  coo_interpolate(skull.min) %>%
   coo_align()  %>%
  coo_slide(id=1) %>% 
  fgProcrustes() %>% 
  efourier(norm=FALSE) %>% 
  PCA()

skull.pca %>% 
  plot_PCA(title = "Dragonfish Skulls")

```

***Figure 3. PCA analysis of Dragonfish Skulls***

## Comparative Analysis

::: {style="text-align: left"}
After acquiring the PCA data, the phylogenetic tree of Lepidoptera was filtered to only include the species that were included in this study.
:::

<center>

```{r, fig.show='hide', display=FALSE}

lep.tree <- ape::read.tree("lep_tree2.tre")

plot(lep.tree,cex=0.1)

```

```{r, echo=FALSE, results='hide'}
lep.tree <- ladderize(lep.tree)
plot(lep.tree,cex=0.1)

lep.tree$tip.label <- gsub("_"," ",lep.tree$tip.label)
basename(names(outs))[1:5]
lep.sp <- read_csv("lep_image_data.csv")
head(lep.sp)
head(lep.sp$identifier)

#
out.data <- tibble(xy.file=basename(names(outs))) %>% 
  mutate(identifier=gsub("XY_|_hindwing|_forewing|.txt","",xy.file)) %>% 
  left_join(lep.sp)

hindwing.pca2 <-  tibble(xy.file=basename(rownames(hindwing.pca$x)),PC1=hindwing.pca$x[,1],PC2=hindwing.pca$x[,2]) %>% 
  left_join(out.data)

forewing.pca2 <-  tibble(xy.file=basename(rownames(forewing.pca$x)),PC1=forewing.pca$x[,1],PC2=forewing.pca$x[,2])%>% 
  left_join(out.data)

```

***Figure 6. Phylogenetic Tree of Lepidoptera***

## Evolutionary Rates

<center>

```{r,echo=FALSE}
drops <- lep.tree$tip.label[!lep.tree$tip.label%in%unique(out.data$species)]

lep.tree2 <- drop.tip(lep.tree,drops)

plot(lep.tree2,cex=0.5)

#PC1s
hind.pc1 <- hindwing.pca2 %>% 
    filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull

names(hind.pc1) <-  hindwing.pca2%>% 
    filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull(species)

fore.pc1 <- forewing.pca2 %>% 
    filter(species%in% lep.tree2$tip.label) %>% 
   group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull(PC1)

names(fore.pc1) <-  forewing.pca2 %>% 
  filter(species%in% lep.tree2$tip.label) %>% 
     group_by(species) %>% 
  summarize(PC1=mean(PC1)) %>% 
  pull(species)

#PC2s
hind.pc2 <- hindwing.pca2 %>% 
    filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>% 
  summarize(PC2=mean(PC2)) %>% 
  pull(PC2)

names(hind.pc2) <-  hindwing.pca2%>% 
    filter(species%in% lep.tree2$tip.label) %>% 
  group_by(species) %>%
  summarize(PC2=mean(PC2)) %>% 
  pull(species)

fore.pc2 <- forewing.pca2 %>% 
    filter(species%in% lep.tree2$tip.label) %>% 
   group_by(species) %>% 
  summarize(PC2=mean(PC2)) %>% 
  pull(PC2)

names(fore.pc2) <-  forewing.pca2 %>% 
  filter(species%in% lep.tree2$tip.label) %>% 
     group_by(species) %>% 
  summarize(PC2=mean(PC2)) %>% 
  pull(species)
```

***Figure 7. Adjusted phylogenetic tree of Lepidoptera included in the study***

```{r, echo=FALSE}
library(phytools)
forePC1.BM<-brownie.lite(lep.tree2,fore.pc1*10)
hindPC1.BM<-brownie.lite(lep.tree2,hind.pc1*10)
forePC2.BM<-brownie.lite(lep.tree2,fore.pc2*10)
hindPC2.BM<-brownie.lite(lep.tree2,hind.pc2*10)

#Evolutionary rates of fore/hindwing
forePC1.BM$sig2.single
hindPC1.BM$sig2.single

forePC2.BM$sig2.single
hindPC2.BM$sig2.single
```

## Shifts in Evolutionary Rates

::: {style="text-align: left"}
The evolutionary shift was then measured to determine if any significant shift in evolutionary rate was present according to the different clades of Lepidoptera under the Brownian model.
:::

<center>

```{r, include= FALSE}
library(RRphylo)
hindPC1.RR <- RRphylo(tree=lep.tree2,y=hind.pc1)
hindPC1.SS<- search.shift(RR=hindPC1.RR,status.type="clade")
forePC1.RR <- RRphylo(tree=lep.tree2,y=fore.pc1)
forePC1.SS<- search.shift(RR=forePC1.RR,status.type="clade")

plot(lep.tree2)
nodelabels(node = as.numeric(rownames(hindPC1.SS$single.clades)),text = rownames(hindPC1.SS$single.clades))

hindPC1.plot <- plotShift(RR=hindPC1.RR,SS=hindPC1.SS)

forePC1.plot <- plotShift(RR=hindPC1.RR,SS=hindPC1.SS)

hindPC1.plot$plotClades()
forePC1.plot$plotClades()
```

```{r, Define funtion, include= FALSE}


plot_SS <- function(tre=NULL,SS=NULL,tax=NULL){
  

  nodes <- as.numeric(rownames(SS$single.clades))
  
  pal <- wes_palette("Zissou1",n=length(nodes))
  sp <- list()
  for(i in nodes){
    sp.i <- extract.clade(tre,i)$tip.label
    
    #print(head(tax))
    sub.names <- lapply(tax,function(x) x[x%in%sp.i]) 
    
    in.clades <- lapply(sub.names,function(x) length(x)>0) 
    all.of.clade <- lapply(sub.names,function(x) all(sapply(sp.i,function(z) z%in%x))) 
    
    high.clade <- names(sub.names)[last(which(all.of.clade==T))]
    all.clades <- names(sub.names)[which(in.clades==T)]
    crown <- ""
    if(high.clade!=last(names(sub.names))) crown <- "crown-"
    
    sub.clades <- NULL
    if(length(grepl("oidea",all.clades))>0) sub.clades <- all.clades[grepl("oidea",all.clades)]

    high.clade2 <- paste0(crown,high.clade,": ",paste0(sub.clades,collapse = "+"))
    sp[[paste0(i)]] <- tibble(n=i,species=sp.i,clade=high.clade2)
    
  }

  
  d <- do.call(rbind,sp)%>% 
    rename(label=species) 
  
  d2<- d %>% rename(clade_name=clade) 
  
  p <- ggtree(tre)+ scale_y_reverse()
  
  p$data <- p$data %>% left_join(d) %>% left_join(tibble(node=nodes,SS$single.clades) %>% mutate(shift=ifelse(rate.difference>0,"+","-")))
  
  p <-  p+geom_tiplab(aes(col=clade),geom="text",size=1.2)+
    geom_cladelab(data=d2,mapping=aes(node=n,col=clade_name,label=clade_name),offset=1,size=1.5)+
    geom_hilight(data=d2,mapping = aes(node = n,fill=clade_name),alpha = 0.01)+
    scale_fill_manual(values = pal)+
    scale_color_manual(values = pal)+
    theme(legend.position = "none")+geom_nodepoint(mapping=aes(subset = shift =="-"), size=5, shape=25,fill='blue',color='blue',alpha=0.7)+
    geom_nodepoint(mapping=aes(subset = shift =="+"), size=5, shape=24, fill='red',color='red',alpha=0.7)
  p <- p+xlim(NA,6)
  res <- tibble(n=nodes,SS$single.clades) %>% left_join(d %>% select(n,clade) %>% unique)
  
  return(list(plot=p,res=res))
  
}

```

```{r, plot shift in evolutionary rates}
tax.names <- readRDS("Lep_classification.RDS")

hindPC1.res <- plot_SS(lep.tree2,hindPC1.SS,tax = tax.names)
forePC1.res <- plot_SS(lep.tree2,forePC1.SS,tax = tax.names)

hindPC1.res$plot
hindPC1.res$res
forePC1.res$plot
forePC1.res$res
```

***Figure 8. Significant evolutionary rate found in the Lepidoptera phylogeny***

## Shape Evolution Correlation

::: {style="text-align: left"}
The PC1 and PC2 data that was underwent phylogenetic independent contrast showed that the PC1 data of forewing and hindwing shape comparison had much more significant correlation compared to the PC2.
:::

### PC1

<center>

```{r, PIC for PC1, echo=FALSE}
hindPC1.pic <- pic(hind.pc1,phy = lep.tree2)
forePC1.pic <- pic(fore.pc1,phy = lep.tree2)


PC1.pic <- tibble(
  hind=hindPC1.pic,
  fore=forePC1.pic
)

PC1.pic %>% 
  ggplot(aes(x=fore,y=hind))+geom_point()+geom_smooth(method="lm")+theme_classic()

summary(lm(hind~fore,PC1.pic))

```

***Figure 9. Phylogenetic Independent Contrast of PC1***

<center>

### PC2

```{r, PIC for PC2, echo=FALSE}
hindPC2.pic <- pic(hind.pc2,phy = lep.tree2)
forePC2.pic <- pic(fore.pc2,phy = lep.tree2)
PC2.pic <- tibble(
  hind=hindPC2.pic,
  fore=forePC2.pic
)

PC2.pic %>% 
  ggplot(aes(x=fore,y=hind))+geom_point()+geom_smooth(method="lm")+ theme_classic() #plot PIC PC2

summary(lm(hind~fore,PC2.pic)) #Find R-square

```

***Figure 10. Phylogenetic Independent Contrast of PC2***

# Discussion

Various studies have been done to analyze the macroevolutionary history of Lepidopteran forewings and hindwings.
One study by @le2019adaptive links morphology with natural selection, but there is little information known about rates of evolution and correlation of wing shapes.
In this study, we aimed to answer some of these questions about evolutionary history.
We first compared the rate of wing-shape evolution between forewings and hindwings.
Evolutionary rates calculated from PC1 for forewings and hindwings differed by 35.56%, whereas evolutionary rates calculated from PC2 differed by 87.11%, indicating a significant shift in evolutionary rates between forewing and hindwing.
Because forewing and hindwing were found to have different evolutionary rates, we examined shift in evolutionary rates of major lineages as it pertains to both hind- and forewing PC1.
Results obtained from both agreed with each other, in that the clade Macroheterocera: Noctuoidea underwent a downward shift in wing-shape evolutionary rate.
Next, to study fore- and hindwing shape correlation, we examined PC1 and PC2 transformed with phylogenetic information for both.
PIC analysis on fore- and hindwing PC1 showed significant correlation (R-squared: 0.4476), whereas PIC analysis using PC2 found no significant correlation (R-squared: 0.00408).

# Author Contributions

Justin - File Formatting, Code, Methods

James -

Chris -

John -

# References
